Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)

### 2024-11-29 paper/paper.qmd
inital

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)"
subtitle: "Analyzing Critical Factors Shaping Mortality and Well-Being Across 193 Nations"
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 30 November 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 193 countries and make the prediction of people from both developed and developing country. Multiple Linear regression is used to deploying the health and socio-economic factors. Predictions of life expectancy of people from developed or developing country is made according to the essential predictors. The finding indicates that the developed countries tend to live longer life as assumed. And we predict that the average age of persons in developed country is 80, while those in underdeveloped countries are 75. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

The data used in this analysis originates from The World Health Organization’s (WHO) and Global Health Observatory (GHO) [@lifeexpectancy]. This data-set related to life expectancy, health factors for 193 countries has been collected from WHO data repository website and its corresponding economic data was collected from United Nation website. Among all categories of health-related factors only those socio-economic factors on the national level were chosen for global scale analysis.

@@

Life Expectancy (`LifeExpectancy`): This variable represents the average number of years a person is expected to live, assuming current mortality conditions persist. It is derived using data from national health records, the World Health Organization (WHO) and Global Health Observatory (GHO). The values are calculated and represented in age between 36.3 to 89.

Percentage Expenditure on Health (`PercentageExpenditure`): This variable reflects the expenditure allocated to health-related expenditures as a percentage of GDP per capita. The data is typically collected from government reports, economic surveys, and global health databases. It indicates the level of personal financial investment in the health sector and is expressed as a percentage.

Schooling (`Schooling`): This variable captures the average number of years of education received by individuals in a country. Data is obtained from educational surveys and censuses conducted by international and national organizations. The values are recorded in years and used to assess the link between education and health outcomes like life expectancy. The data range from 0 to 20.7

Status (`Status`): This is a categorical variable that classifies countries as "Developed" or "Developing," based on various socio-economic and health indicators accepted globally. It provides context for comparing life expectancy across different stages of national development.

Continent(`Continent`): This is a categorical variable that classifies the geographic location of the countries. It is detected by the name of the country using rpackage `countrycode`. The data is stored as the name of the continent(‘Africa’, ‘Oceania’, ‘Asia’, ‘Europe’, ‘North America’, ‘South America’)

Total expenditure (`TotalExpenditure`): This is a numerical variable that shows the general government expenditure on health as a percentage of total government expenditure. The number drop of the percentage and shows as the number between 0 to 100, where higher values indicate a greater share of government spending allocated to health.

Income composition of resource (`IncomeComposition`): This is a numerical variable representing the income component of the Human Development Index (HDI). It reflects the contribution of income to human development, with values ranging from 0 to 1, where higher values signify a greater income contribution to overall development.

@@

The outcome variable is `LifeExpendency`. This is the primary dependent variable that the model is designed to predict. It represents the average number of years a person is expected to live, under the condition that current mortality conditions persist. The model seeks to identify the variables affecting this average. To examine the differences in life expectancy under the same variables, we separate the data into developed and developing countries, where developed country are generally believed to have higher life expectancy. @fig-expectancy displays density of the historical data of life expectancy across 195 countries from 2000 to 2015, comparing developing to developed countries.The average age of most developed country is approximately 80, while the majority developing countries typically have average life expectancy around mid-seventies.

```{r}
#| label: fig-expectancy
#| fig-cap: This histogram shows the distribution of the percentage of life expectancy dividing by the status of countries. The graph of developed countries shows narrower spread with density curve located on the rightside of the x-axis. The graph of developing countries performs a wider spread while the highest density position at mid-seventies.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot of historical 193 countries life expectancy dividing by country status. 
ggplot(cleaned_expectancy, aes(x = LifeExpectancy)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 70, fill = "salmon", color = "pink", alpha = 0.9) +
  geom_density(color = "royalblue", fill = "lightblue", alpha = 0.6) +
  labs(
    title = "Distribution of Life Expectancy by Status",
    x = "Life Expectancy", y = "Density"
  ) +
  theme_minimal() +
  facet_wrap(~Status)

@@

1. **Status(`Status`)**:

2. **Continent(`Continent`)**:

@@

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

#analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

#analysis_data |> 
  #ggplot(aes(x = width, y = length)) +
  #geom_point(alpha = 0.8) +
  #theme_minimal() +
  #labs(x = "Wing width (mm)",
      # y = "Wing length (mm)")
```

# Model
To predict the outcome of the life expectancy of people from different regions, we developed tow linear regression models: one for the developed country and one for developing country. These models aim to estimate the life expectancy of people under different environment.

@@

>>>>

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)"
subtitle: "Analyzing Critical Factors Shaping Mortality and Well-Being Across 155 Nations"
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 30 November 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 155 countries and make the prediction of people from both developed and developing country. Multiple Linear regression is used to deploying the health and socio-economic factors. Predictions of life expectancy of people from developed or developing country is made according to the essential predictors. The finding indicates that the developed countries tend to live longer life as assumed. And we predict that the average age of persons in developed country is 80, while those in underdeveloped countries are 75. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

The data used in this analysis originates from The World Health Organization’s (WHO) and Global Health Observatory (GHO) [@lifeexpectancy]. This data-set related to life expectancy, health factors for 155 countries has been collected from WHO data repository website and its corresponding economic data was collected from United Nation website. Among all categories of health-related factors only those socio-economic factors on the national level were chosen for global scale analysis.

@@

**Life Expectancy (`LifeExpectancy`)**: This variable represents the average number of years a person is expected to live, assuming current mortality conditions persist. It is derived using data from national health records, the World Health Organization (WHO) and Global Health Observatory (GHO). The values are calculated and represented in age between 36.3 to 89.

**Percentage Expenditure on Health (`PercentageExpenditure`)**: This variable reflects the expenditure allocated to health-related expenditures as a percentage of GDP per capita. The data is typically collected from government reports, economic surveys, and global health databases. It indicates the level of personal financial investment in the health sector and is expressed as a percentage.

**Schooling (`Schooling`)**: This variable captures the average number of years of education received by individuals in a country. Data is obtained from educational surveys and censuses conducted by international and national organizations. The values are recorded in years and used to assess the link between education and health outcomes like life expectancy. The data range from 0 to 20.7

**Status (`Status`)**: This is a categorical variable that classifies countries as "Developed" or "Developing," based on various socio-economic and health indicators accepted globally. It provides context for comparing life expectancy across different stages of national development.

**Continent(`Continent`)**: This is a categorical variable that classifies the geographic location of the countries. It is detected by the name of the country using rpackage `countrycode`. The data is stored as the name of the continent(‘Africa’, ‘Oceania’, ‘Asia’, ‘Europe’, ‘North America’, ‘South America’)

**Total expenditure (`TotalExpenditure`)**: This is a numerical variable that shows the general government expenditure on health as a percentage of total government expenditure. The number drop of the percentage and shows as the number between 0 to 100, where higher values indicate a greater share of government spending allocated to health.

**Income composition of resource (`IncomeComposition`)**: This is a numerical variable representing the income component of the Human Development Index (HDI). It reflects the contribution of income to human development, with values ranging from 0 to 1, where higher values signify a greater income contribution to overall development.

@@

The outcome variable is **`LifeExpendency`**. This is the primary dependent variable that the model is designed to predict. It represents the average number of years a person is expected to live, under the condition that current mortality conditions persist. The model seeks to identify the variables affecting this average. To examine the differences in life expectancy under the same variables, we separate the data into developed and developing countries, where developed country are generally believed to have higher life expectancy. @fig-expectancy displays density of the historical data of life expectancy across 195 countries from 2000 to 2015, comparing developing to developed countries.The average age of most developed country is approximately 80, while the majority developing countries typically have average life expectancy around mid-seventies.

```{r}
#| label: fig-expectancy
#| fig-cap: "This histogram shows the distribution of the percentage of life expectancy dividing by the status of countries. The graph of developed countries shows narrower spread with density curve located on the rightside of the x-axis. The graph of developing countries performs a wider spread while the highest density position at mid-seventies."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# create plot of historical 155 countries life expectancy dividing by country status. 
ggplot(cleaned_expectancy, aes(x = LifeExpectancy)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 70, fill = "salmon", color = "pink", alpha = 0.9) +
  geom_density(color = "royalblue", fill = "lightblue", alpha = 0.6) +
  labs(
    title = "Distribution of Life Expectancy by Status",
    x = "Life Expectancy", y = "Density"
  ) +
  theme_minimal() +
  facet_wrap(~Status)

@@

1.  **Status(`Status`)**: The status of a country is the key factor influencing the life expectancy. We divide the life expectancy by the status of a country as it is believed that developed country generally have higher life expectancy due to better healthcare access, sanitation, and living standards, while developing countries face greater health challenges. This variable provides context for comparing life expectancy between different levels of country development. @fig-statusdistribution includes 28 developed countries and 127 developing countries. The graph on left shows that we've got much greater representation of developing nations across the dataset. Comparing both graph, we still could tell that the dataset contains significantly more data for developing countries than for developed countries, even relative to the difference in the number of countries themselves.

2.  **Continent(`Continent`)**: This is the categorical variable showing the geographic information of each country. Continent usually

@@

```{r}
#| label: fig-statusdistribution
#| fig-cap: "Distribution of country status, showing 28 dveloped countries and 127 developing countries, with a substantial difference in the number of data points. From 2000 to 2015, notably, the number of data entries for developing countries far exceeds that of developed countries, reflecting the greater representation of developing nations across the dataset."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Count the number of unique countries for each Status
country_counts <- cleaned_expectancy %>%
  distinct(Country, Status) %>%  # Ensure each country is only counted once
  count(Status)  # Count the number of countries in each Status category

#Plot1: create bar plot of count of status
numberrows <- ggplot(cleaned_expectancy, aes(x = Status, fill = Status)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    title = "Distribution of Count of Developed and Developing Countries",
    x = "Status (Developed vs Developing)",
    y = "Count"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(legend.position = "none",
  plot.title = element_text(size = 7))

# Plot2: Create a bar plot to show the distribution of countries by Status
numbercountry <- ggplot(country_counts, aes(x = Status, y = n, fill = Status)) +
  geom_bar(stat = "identity") +  # Use 'identity' to directly use the count values
  geom_text(aes(label = n), vjust = -0.5) +  # Add count labels on top of bars
  labs(
    title = "Distribution of Countries by Status (Developed vs Developing)",
    x = "Status (Developed vs Developing)",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(
    legend.position = "none", 
  plot.title = element_text(size = 7))

# Arrange the plots side by side
grid.arrange(numberrows, numbercountry, ncol = 2)
```

```{r}
#| label: fig-continent
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. "
#| echo: false
#| eval: true
#| warning: false
#| message: false

#Calculate the number of unique countries per continent
continent_count <- cleaned_expectancy %>%
  group_by(Continent) %>%
  summarise(num_countries = n_distinct(Country))

#Plot1: Distribution of number of data per continent
#Create the bar plot showing the number of data per continent
continentrow <- ggplot(cleaned_expectancy, aes(x = Continent)) +
  geom_bar(fill = "orange", color = "white") +
  labs(
    title = "Distribution of Count by Continent",
    x = "Continent",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12)
  ) +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5, size = 2.5)

# Plot2: The distribution of number of countries each continent
#Create the bar plot showing the number of unique countries per continent
continentnumber <-ggplot(continent_count, aes(x = Continent, y = num_countries)) +
  geom_bar(stat = "identity", fill = "#FFA07A", color = "white") +
  labs(
    title = "Number of Countries by Continent",
    x = "Continent",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10)
  ) +
  geom_text(aes(label = num_countries), vjust = -0.5, size = 2.5)

# Arrange the plots side by side
grid.arrange(continentnumber, continentrow, ncol = 2)
```

# Model

To predict the outcome of the life expectancy of people from different regions, we developed tow linear regression models: one for the developed country and one for developing country. These models aim to estimate the life expectancy of people under different environment.

@@

1. check different continent has its different life expectancy. 
2. compare the develop country with 
3. check the shcooling year of develop countries to developing country, showing a relationship between shcooling with the life expenctancy. 
4.
